Effects of below-ground sink limitation on growth and carbon balance of Eucalyptus seedlings

### 2015-07-22 manuscript/manuscript_pve.Rmd
changes to bibtex CO2
edits to manuscript

small edits to figures and table dfrs

---
title: "Effects of below-ground space limitation on performance of Eucalyptus seedlings:  Does photosynthesis really control growth?"
author: ""
date: ""
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: tree-physiology.csl
bibliography: references.bib
---

@@

Interpreting limitations to plant growth requires understanding of the balance between carbon (C) source and sink activity in order to assess C allocation and biomass partitioning. This study used manipulations of soil volume to test how growth is coupled to physiology, allocation, and sink activity in *Eucalyptus tereticornis* seedlings. We grew seedlings in a large range of container sizes and planted containers flush to the soil alongside naturally sown seedlings (free). Reduced soil volume was expected to induce rapid negative effects on growth and physiology compared to free seedlings. It was hypothesized that soil volume effect would be largest in the smallest containers, mediated by a down-regulation of photosynthesis (A). The accumulation of leaf non-structural carbohydrates, resulting from reduced sink strengh, was expected to correleltae to reductions in photosynthetic capacity. We observed a negative container effect on aboveground growth soon after the experiment started. Although growth was consistently different across soil volumes mass, partitioning to leaves, stems and roots was conserved after 120 days. Photosynthetic capacity was also significantly reduced in containers, and was related to both leaf nitrogen content and starch accumulation. We developed a seedling growth model that utilized leaf A rates to allocate daily C uptake towards mass growth of stems, leaves and roots. We then asked whether the observed reductions in A explained the observed differences in seedling biomass. We found that although belowground sink limitation resulted in the down regulation of A, these reductions were not significant enough to explain observed growth responses. Thus, as photosynthesis and growth were not coordinated an excess pool of non-biomass C resulted in seedlings with soil volume restriction. This research highlights the need to further utilize mass balance approaches when evaluating plant C allocation and confirms that A and growth are not always directly related.

@@

Understanding plant growth requires knowledge of the mass balance that must be achieved between C uptake and subsequent allocation to growth, storage, and respiration. As woody plants have highly integrated systems of competing carbohydrate sinks [@kozlowski1992carbohydrate], growth should principally depend on the allocation of photosynthate among different tissues and organs. At long enough time scales A and growth must be correlated in respect to mass balance, however, at shorter temporal scales growth is not necessarily limited by the availability of recent photosynthate. This has led to the current debate over how strongly plant growth is controlled by either source of sink activity. Consequently, plant growth cannot be simply determined by rates of A making it complex to understand and challenging to model [@fourcaud2008plant]. Despite a wealth of studies large uncertainties still remain regarding the coordination of C supply, via A, and growth of woody species.

\
In woody species, the coordination of A and growth has been studied with manipulations of C source activity. Examples included elevated CO~2~ experiments, for example FACE [reviewed in @ainsworth2005have], and partial defoliation experiments. Elevated CO~2~ has been shown to increase A rates [@drake1997more; @ainsworth2007response] and across four FACE experiments this resulted in a conserved increase in forest production [@norby2005forest]. Evidence from elevated CO~2~ experiments, however, also reveals that the growth response tends to be much smaller than the photosynthetic enhancement [@kirschbaum2011does]. In defoliation experiments, compensatory increases in A are commonly shown yet are attributed to variable mechanisms, including reduction in end product inhibition [@iglesias2002regulation; @zhou2003changes; @handa2005test], enhanced biochemical activity [@ovaska1993b;@layne1995end], increased stomatal conductance [@layne1995end], leaf nutrient status [@turnbull2007increased], and regulatory sugar signaling [@eyles2013whole]. However, increases in A did not always produce increased growth due to reductions in meristem sink strength [@palacio2012fast], C limitation to mycorrhizal colonization [@markkola2004defoliation], or an overall decrease in whole plant C gain [@ovaska1993a]. These manipulations of C source activity expose unresolved issues with how changes in A do not always infer similar responses in growth.

\
Alternatively, manipulating plant tissue C sinks is often used to investigate the correlation of A and growth. This is because metabolic signaling networks, relaying information on C and N status of different tissues, can down regulate photosynthetic activity [@paul2001sink]. If this sink inhibition of A occurs then a close coordination between declines in A and growth should be expected. Whether this photosynthetic down regulation is evident in woody species has been tested through fruit removal, girdling, and low temperatures. In these studies, down regulation of A was frequently correlated to carbohydrate accumulation resulting from reduced tissue sink strength [@iglesias2002regulation;@hoch2002altitudinal; @urban2007girdling; @haouari2013fruit]. However, reductions in A were also attributed to biochemical limitations prior to carbohydrate accumulation [@nebauer2011photosynthesis], irreversible photo-oxidative damage [@duan2008photosynthetic], and stomatal limitation [@li2005photosynthesis]. These mixed results are not surprising as we still know little about the pathways in which plants achieve balance between assimilation, storage, and growth across temporal scales [@smith2007coordination]. As these manipulations likely impact source as well as sink activity simultaneously, affect water transport, are very extreme, or are specific to large annual fruiting sinks, they tell us little about source-sink coordination in normal field-grown woody species.

\
If source or sink manipulations do not lead to equivalent changes in growth then understanding the fate of assimilated C becomes even more crucial. As programmed plasticity in C allocation naturally occur as plants growth and develop, this must be accounted for when evaluating induced variation in growth [@wright2002interpreting]. Thus, when C allocation is altered by a treatment, it should be shown whether these allocation patterns differ from common sized untreated plants [@reich2002root;@poorter2012biomass]. Only then can we test if functional balance preservation or optimal foraging occurs from changes in C uptake. For example, seedlings in containers may shift allocation to leaves if root restriction occurs or alter fine root morphology to increase nutrient uptake. In woody species, shifts in allometric relationships of leaves, stems and roots can reveal if growth is ontogenetically constrained or actively adjusted when A is affected. This then allows additional pools of C, such as root exudation or changes in tissue respiration, to be accounted for and mass balance to be achieved.

@@

1). The manipulations of container size were expected to induce a belowground sink limitation compared to free seedlings. We hypothesized that declines in seedling grow would be largest in the smallest containers, mediated by a down-regulation of photosynthesis.

2). Reduced sink strength was expected to lead to accumulation of leaf non-structural carbohydrates, which is known to inhibit A We therefore expected a correlation between carbohydrate accumulation and photosynthetic capacity as a function of soil volume.

3). Last, the growth model was expected to find agreement between simulated and observed seedling mass, through direct correlation of the effects of limiting soil volume on A.

@@

## Growth and morphology metrics
Seedlings were planted on January 21^st^ 2013 and stem height, diameter at 15 cm and leaf count were measured weekly thereafter. Once the growth rate of individual plants had significantly declined a full biomass harvest was completed (May 21^st^ 2013). Dry mass of leaves, stems, roots and cumulative leaf area (LI-3100C Area Meter; LI-COR, Lincoln, NE, USA) was measured for each seedling. Mean individual leaf area for each harvested seedling was calculated by dividing cumulative leaf area by total leaf count of only fully expanded leaves. This value was then used to interpolate cumulative leaf area through time with weekly leaf counts. Root mass was collected by passing soil from each container through a 1 mm sieve, washing, separating into fine and coarse roots (<2 mm and >2 mm diameter, respectively) and then drying to a constant mass. Roots from the free seedlings were collected by excavating each 1 m^2^ subplot to 0.5 m depth.  25 g fresh weight subsamples of washed fine roots were analyzed, using WhinoRhizo software (Regent Instruments Inc.), for specific root length (SRL, m g^-1^).

@@

\
Leaf TNC concentration was analyzed using a total starch assay kit (Megazyme International 303 Ireland Ltd., Wicklow, Ireland) and includes the starch (mg g^-1^) and soluble sugar (mg g^-1^) concentrations. Starch was quantified using a thermostable $\alpha$-amylase and amyloglucosidase assay [@McCleary_starch] and soluble sugars were determined following the anthrone method [@ebell1969variation]. Complete methods of the TNC assay are described in [@mitchell2013drought]. Specific leaf area (SLA, m^2^ kg^-1^), for leaves sampled during gas exchange campaigns, was then calculated by first subtracting the TNC content from individual dry leaf mass before dividing leaf area by leaf mass.

@@

\
We then utilized this model to test the hypothesis that the effects of belowground sink limitation on rates of leaf A where sufficient to accurately predict overall seedling biomass production after 120 days. Each model run utilized changes in A and leaf mass fraction (LMF), with published or local values of stem and root respiration rates, to generate total seedling mass and leaf area after 120 days.  Cumulative net leaf carbon gain for each treatment was equal to the sum of each value of C~day,i~ over 120 days and final seedling carbon was assumed to equal half of the final mass for both modeled and observed seedlings. First, a default model was optimized to produce a final LMF that correctly predicted both the final leaf mass and total biomass of the harvested free seedling controls (M~0~). This optimized LMF was then applied to model runs with treatment specific C~day~ to determine if changes in leaf A alone could accurately predict biomass (M~1~). Next, model sensitivity to different C allocation scenarios, including non-optimized treatment specific LMF and up regulation of non-leaf tissue respiration by 50 % of default values (M~2,3~, respectively), was used to improve predictions of initial model simulations from measured harvest biomass. For all cases, seedling production was compared between model output and harvested seedlings with treatment specific mean daily carbon assimilation by first scaling values to the free seedling control.

##Data analysis
Differences in experimental parameters with soil volume were analysed by one-way analysis of variance (ANOVA) in R with individual containers as random effects and soil volume as a categorical fixed effect. Tukey’s post-hoc tests were performed in conjunction with ANOVA to determine which specific paired comparisons among soil volume treatments were different. Mixed model ANOVAs of A~max~ and leaf chemistry were performed using the 'nlme' package [@nlme] in R and R^2^ values of mixed models were computed as in [@nakagawa2013general]. Tests of allometric relationships between biomass components were implemented using major axis regression in the 'smatr' package in R [@warton2012smatr]. Results were considered significant at P&leq;0.05.

@@

## Growth and morphology metrics
In this field study, colder temperatures and reductions in cumulative PPFD per day (Fig. `r airvarsfig`) most likely lead to the reduced growth in the free seedlings in the final weeks of the experiment (Fig. `r allomfig`).  Combined with severe growth reductions in the smallest container volumes the experiment was chosen to be harvested after 120 days. Over this duration height, diameter, and leaf area diverged between container volumes (Fig. `r allomfig`).  First, seedling leaf area significantly diverged between soil volumes (P<0.026) during the 5^th^ week of the experiment. Following this period both height (8^th^} week) and then diameter (9^th^ week) significantly deviated across soil volumes (P<0.002 & 0.001, respectively).  Negative growth effects then manifested as severely reduced height gain and declining leaf area through time with small soil volumes across the final two months of the experiment. Seedlings maintained diameter growth throughout the experiment, although marginal with smaller soil volumes in the final month. Final seedling height significantly increased with increasing soil volume (P<0.001).  Increases in both final stem diameter (P<0.001) and cumulative leaf area (both P<0.001) were found with increasing soil volume and these differences were driven mainly by the largest container and the free seedling.

\
Total seedling biomass at harvest was significantly different across container volumes (P<0.001) and with free seedlings (P<0.001, Table. 1). We analyzed the relationship between biomass growth with each fold increase in soil volume and found an increase of 34% with a doubling of pot size, consistent with the meta-analysis of Poorter et al. [-@poorter2012pot]. Additionally, plant biomass was highly correlated with total leaf area across all treatments (R^2^ = 0.97, P<0.001). Differences in biomass partitioning to leaves, stems, and roots were not different across soil volumes after variation in seedling biomass within treatments was factored in the analysis (Fig.`r allocfig`a,b). Across all treatments, the final harvested root:shoot was conserved in these seedlings, with a slightly higher shoot than root mass on average ($\overline{x}$=0.904, Fig. `r allomfig`c).

\
Overall, SRL was higher in seedlings in containers than free seedlings but not significantly in every soil volume treatment   (Table. 1). Over the duration of the experiment SLA was higher in free seedlings but was not different across containers sizes (Table. 1, P<0.001) and this pattern was evident in the first gas exchange measurement campaign (P<0.001).

## Leaf chemistry
Leaf N % was significantly higher in free seedlings and the largest container volume at the onset of gas exchange measurements (6th week, P<0.001).  Over the remaining duration of the experiment the smallest container volume had a significant reduction in leaf N % compared to other soil volumes, while free seedlings maintained the highest leaf N % (Table. 1, P<0.001).  Leaf starch content in the smallest container was double that of free seedlings (P=0.039), while leaf soluble sugars did not differ across treatments throughout the experiment (Table. 1).  Differences in leaf starch between the free seedling and the smallest container were also evident during the first gas exchange campaign (P=0.0013).

## Gas exchange and photosynthetic parameters
A~sat~ and A~max~ were both significantly higher in the largest container volume and the free seedling at the first measurement campaign (both P<0.001). Across all measurement campaigns A~sat~ (Fig. `r asatfig`) and A~max~ (Table 2.) were consistently higher in free seedlings than in containers (both P<0.001). The interaction between photosynthetic capacity, leaf starch, and leaf N on a mass basis was marginally significant (P=0.0584) but A~max~ was highly correlated to both leaf N content and leaf starch (both P<0.001). Across all measurement campaign A~max~ was higher when foliar N was also higher, usually associated with low levels of leaf starch (Fig. `r achemfig`a). A~max~  was also lower when leaf starch was high as higher leaf N often did not coincide with high leaf starch (Fig. `r achemfig`b)

\
The photosynthetic parameters J~max~ and Vc~max~ were not different within treatments at the beginning and end of gas exchange campaigns, therefore the parameter means per treatment are reported here (Table. 2).  Overall, both J~max~ and Vc~max~ were significantly higher in free seedlings with little variation between soil volume treatments (P=0.0012 & 0.0021, respectively). Leaf dark respiration rates were not significantly different across soil volumes (Table. 2). The g~1~ parameter, generated for each seedling from the Medlyn et al [-@medlyn2012reconciling] optimal stomatal conductance model, was lowest in the free seedling and was marginally different across soil volume treatments (Table. 2). Predicted values of g~s~, using the g~1~ parameter, where highly correlated with observed values (R^2^= .74, P<0.001, data not shown).

\
Neither $\Psi$~pd~ nor $\Psi$~l~ were different across treatments, with mean values of -0.27 and -1.2 mPa across all seedlings, respectively. Although P=0.001), the mean rates for all seedlings were high at 0.37 mol H~2~0 m^-2^ s^-1^ and did not decline significantly across the experiment duration. Additionally, leaf $\delta$^13^C at final harvest was not different across treatments (Table. 1). Combined these indices provide strong evidence that water stress was not apparent on these well-watered seedlings throughout the experiment. Soil N % at harvest was not different across soil volumes ($\bar{x}$=04.5 %) and decreased approximately 3 % across all containers over the experiment duration. This indicates that nutrient leaching from free seedlings or from draining of containers following natural rainfall events did not differ between treatments.

@@

# Discussion
This study utilized a simple but novel field design to manipulate belowground sink limitation and physically restrict *Eucalyptus tereticornis* seedling production.  We then addressed questions regarding the coordination of A and growth by complementing empirical results with modelling approaches. We found that reductions in leaf A rates were not sufficient to explain observed reductions in total plant biomass production. We thus encourage the utilization of mass balance approaches and provide direction for future studies when testing factors that control of plant growth under environmental change.

## Soil volume restiction compared to field plants
As soon as seedlings became established both height and diameter growth were negatively affected as a function of available soil volume when compared to free seedlings. These growth reductions were expected, as the impedance of root growth can cause reductions overall plant growth and activity [@mcconnaughay1991physical; @young1997mechanical]. It has been shown that roots undergoing difficult conditions may send inhibitory signals to the shoots that affect g~s~, cell expansion, cell division and the rate of leaf appearance [@passioura2002soil].  Here, this was evident in a large divergence in leaf area between seedlings in containers and free seedlings through time, with the eventual cessation of new leaf growth in seedlings in small containers.

\
Decreases in A~sat~ paralleled the reductions in allometric growth parameters of seedlings in containers.  This initially suggests a tight link between A and growth, consistent with our hypothesis that growth declines would be mediated by the down regulation of A. However, there are several possible mechanisms that can explain reduced A in small pots, including nutrient content, water or reduced sink strength [@poorter2012pot].  It was therefore necessary to examine each of these factors to distinguish if the induced belowground sink limitation actually triggered photosynthetic down regulation.

\
With high rates of g~s~, non-limiting leaf water potential and consistent leaf $delta$^13^C across soil volume treatments there was little evidence of water supply inducing a source limitation to A. This finding is consistent with other container size studies without drought treatments.  For example, reduced A~max~ in cotton seedlings grown at elevated CO^2^ was attributed to sink-limited feedback inhibition from inadequate rooting volume, not decreased g~s~ [@thomas1991root]. Additionally, severe reductions in A in coffee plants was not attributed to impacts of container size on leaf water potentials or g~s~ [@ronchi2006growth]. In these well watered seedlings it was likely that reductions in A were instead the result of a limited soil nutrient resource pool on belowground sink strength.

\
Although the soil N pool declined evenly across all treatments leaf N was lowest in the smallest containers, suggesting that sink limitation was the greatest in these containers. This makes sense as small containers may reduce N uptake, either from physical root restriction or decreased supply, which affects growth, Rubisco limitation, sugar metabolism, and carbohydrate partitioning between source and sink tissues [@stitt1991rising; @hermans2006plants]. Mycorrhizal colonization could also have been affected by the container effect. Whether unrestricted mycorrhizal recruitment facilitated the increase in leaf N uptake to leaves in free seedlings is unknown. Regardless of whether the sink limitation occurred from root or mycorrhizal restriction, the stark contrast between leaf N between free seedlings and seedlings in containers was significantly correlated to reductions in A.

\
As both rooting space and resources were finite in containers, the limited ability of these seedlings to increase the capacity of the belowground C sink was hypothesized to result in the buildup of C assimilate in the leaves. The feedback inhibition of A from starch accumulation has been proposed, yet it is still not known whether there is a starch threshold that triggers the down-regulation process (Nebauer et al. 2011).  Here, declines in A~max~ where correlated with higher starch content throughout the experiment.  This agrees with findings from Equiza et a. (2006) where photosynthetic downregulation from reduced sink strength showed a greater correlation with starch than with sugar content in a deciduous conifer.  As starch content in the smallest containers was nearly double that of free seedlings this suggests the response of A to sink inhibition was regulated by this accumulation, as hypothesized.

>>>>

---
title: "Effects of below-ground sink limitation on growth and carbon balance of Eucalyptus seedlings"
author: ""
date: ""
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: tree-physiology.csl
bibliography: references.bib
---

@@

Interpreting limitations to plant growth requires understanding of the balance between carbon (C) source and sink activity in order to assess C allocation and biomass partitioning. This study used manipulations of soil volume to test how growth is coupled to physiology, allocation, and sink activity in *Eucalyptus tereticornis* seedlings. We grew seedlings in a large range of container sizes and planted containers flush to the soil alongside naturally sown seedlings (free). Reduced soil volume was expected to induce rapid negative effects on growth and physiology compared to free seedlings. It was hypothesized that soil volume effect would be largest in the smallest containers, negatively impacting mass partitioning belowground. The accumulation of leaf non-structural carbohydrates, resulting from reduced sink strengh, was expected to correleltae to reductions in photosynthetic capacity. We observed a negative container effect on aboveground growth soon after the experiment started. Although growth was consistently different across soil volumes mass, partitioning to leaves, stems and roots was conserved after 120 days. Photosynthetic capacity was also significantly reduced in containers, and was related to both leaf nitrogen content and starch accumulation. We developed a seedling growth model that utilized leaf photosynthesis (A) rates to allocate daily C uptake towards mass growth of stems, leaves and roots. We then asked whether the observed reductions in A explained the observed differences in seedling biomass. We found that although belowground sink limitation resulted in the down regulation of A, these reductions were not significant enough to explain observed growth responses. Thus, as A and growth were not coordinated an excess pool of non-biomass C resulted in seedlings with soil volume restriction. This research highlights the need to further utilize mass balance approaches when evaluating plant C allocation and confirms that A and growth are not always directly related.

@@

Understanding plant growth requires knowledge of the mass balance that must be achieved between C uptake and subsequent allocation to growth, storage, and respiration. As woody plants have highly integrated systems of competing carbohydrate sinks [@kozlowski1992carbohydrate], growth should principally depend on the allocation of photosynthate among different tissues and organs. At long enough time scales photosynthesis (A) and growth must be correlated to maintain mass balance, however, at shorter temporal scales growth is not necessarily limited by the availability of recent photosynthate. This has led to the current debate over how strongly plant growth is controlled by either source of sink activity. Consequently, plant growth cannot always be simply determined by the photosynthesis rate, making it complex to understand and challenging to model [@fourcaud2008plant]. Despite a wealth of studies large uncertainties still remain regarding the coordination of C supply and growth of woody species.

\
In woody species, the coordination of A and growth has been studied with manipulations of C source activity. Examples included elevated CO~2~ experiments, for example FACE [reviewed in @ainsworth2005have], and partial defoliation experiments. Elevated CO~2~ has been shown to increase A rates [@drake1997more; @ainsworth2007response] and across four FACE experiments this resulted in a conserved increase in forest biomass production [@norby2005forest]. Evidence from elevated CO~2~ experiments, however, also reveals that even with average photosynthetic enhancement of over 30 % the growth rate only increases by around 10 %  [@kirschbaum2011does]. In defoliation experiments, increases in A of the remaining foliage are commonly shown yet are attributed to variable mechanisms, including reduction in end product inhibition [@iglesias2002regulation; @zhou2003changes; @handa2005test], enhanced biochemical activity [@ovaska1993b;@layne1995end], increased stomatal conductance [@layne1995end], leaf nutrient status [@turnbull2007increased], and regulatory sugar signaling [@eyles2013whole]. However, increases in A did not always produce increased growth due to reductions in meristem sink strength [@palacio2012fast], C limitation to mycorrhizal colonization [@markkola2004defoliation], or an overall decrease in whole plant C gain [@ovaska1993a]. These manipulations of C source activity expose unresolved issues with how changes in A do not always infer similar responses in growth.

\
Alternatively, manipulating plant tissue C sinks is often used to investigate the correlation of A and growth. This is because metabolic signaling networks, relaying information on C and N status of different tissues, can down regulate photosynthetic activity [@paul2001sink]. If this sink inhibition of A occurs then a close coordination between declines in A and growth should be expected. Whether this photosynthetic down regulation is evident in woody species has been tested through fruit removal, girdling, and low temperatures. In these studies, down regulation of A was frequently correlated to carbohydrate accumulation resulting from reduced tissue sink strength [@iglesias2002regulation;@hoch2002altitudinal; @urban2007girdling; @haouari2013fruit]. However, reductions in A were also attributed to biochemical limitations prior to carbohydrate accumulation [@nebauer2011photosynthesis], irreversible photo-oxidative damage [@duan2008photosynthetic], and stomatal limitation [@li2005photosynthesis]. These mixed results are not surprising as we still know little about the pathways in which plants achieve balance between assimilation, storage, and growth across temporal scales [@smith2007coordination]. As these manipulations likely impact source as well as sink activity simultaneously, affect water transport, are very extreme, or are specific to large annual fruiting sinks, they tell us little about source-sink coordination in typical field conditions for woody species.

@@

\
Our hypotheses were as follows:
\
1). The manipulations of container size were expected to induce a belowground sink limitation compared to free seedlings. We hypothesized that declines in seedling growth would be largest in the smallest containers.

2). As the finite pool of rooting volume and soil nutrients will decline faster in small containers we expected decreases in partitioning to fine root mass with decreasing container size.

3). Reduced sink strength was expected to lead to accumulation of leaf non-structural carbohydrates, which is known to inhibit A. We therefore expected a correlation between carbohydrate accumulation and photosynthetic capacity as a function of soil volume.

4). Last, the growth model was expected to find agreement between observed seedlings mass and mass predicted from a simple carbon balance model taking into account measured rates of photosynthesis.

@@

## Growth and morphology metrics
Seedlings were planted on January 21^st^ 2013 and stem height, diameter at 15 cm and leaf count were measured weekly thereafter. Once the growth rate of individual plants had significantly declined a full biomass harvest was completed (May 21^st^ 2013). Dry mass of leaves, stems, roots and cumulative leaf area (LI-3100C Area Meter; LI-COR, Lincoln, NE, USA) was measured for each seedling. Mean individual leaf area for each harvested seedling was calculated by dividing cumulative leaf area by total leaf count of only fully expanded leaves. This value was then used to interpolate cumulative leaf area through time with weekly leaf counts. Root mass was collected by passing soil from each container through a 1 mm sieve, washing, separating into fine and coarse roots (<2 mm and >2 mm diameter, respectively) and then drying to a constant mass. Roots from the free seedlings were collected by excavating each 1 m^2^ subplot to 0.5 m depth.  25 g fresh weight subsamples of washed fine roots were analyzed, using WhinoRhizo software (Regent Instruments Inc., Quebec, QC, Canada), for specific root length (SRL, m g^-1^).

@@

\
Leaf TNC concentration was analyzed using a total starch assay kit (Megazyme International, Wicklow, Ireland) and includes the starch (mg g^-1^) and soluble sugar (mg g^-1^) concentrations. Starch was quantified using a thermostable $\alpha$-amylase and amyloglucosidase assay [@McCleary_starch] and soluble sugars were determined following the anthrone method [@ebell1969variation]. Complete methods of the TNC assay are described in [@mitchell2013drought]. Specific leaf area (SLA, m^2^ kg^-1^), for leaves sampled during gas exchange campaigns, was then calculated by first subtracting the TNC content from individual dry leaf mass before dividing leaf area by leaf mass.

@@

\
We then utilized this model to test the hypothesis that the effects of belowground sink limitation on rates of leaf A where sufficient to accurately predict overall seedling biomass production after 120 days. Each model run utilized changes in A and leaf mass fraction (LMF), with published or local values of stem and root respiration rates, to generate total seedling mass and leaf area after 120 days.  Cumulative net leaf C gain for each treatment was equal to the sum of each value of C~day,i~ over 120 days and final seedling C was assumed to equal half of the final mass for both modeled and observed seedlings. First, a default model was optimized to produce a final LMF that correctly predicted both the final leaf mass and total biomass of the harvested free seedling controls (M~0~). This optimized LMF was then applied to model runs with treatment specific C~day~ to determine if changes in leaf A alone could accurately predict biomass (M~1~). Next, model sensitivity to different C allocation scenarios, including non-optimized treatment specific LMF and up regulation of non-leaf tissue respiration by 50 % of default values (M~2,3~, respectively), was used to improve predictions of initial model simulations from measured harvest biomass. For all cases, seedling production was compared between model output and harvested seedlings with treatment specific mean daily C assimilation by first scaling values to the free seedling control.

##Data analysis
Differences in experimental parameters with soil volume were analysed by one-way analysis of variance (ANOVA) in R with individual containers as random effects and soil volume as a categorical fixed effect. Tukey's post-hoc tests were performed in conjunction with ANOVA to determine which specific paired comparisons among soil volume treatments were different. Mixed model ANOVAs of A~max~ and leaf chemistry were performed using the 'nlme' package [@nlme] in R and R^2^ values of mixed models were computed as in [@nakagawa2013general]. Tests of allometric relationships between biomass components were implemented using major axis regression in the 'smatr' package in R [@warton2012smatr]. Results were considered significant at P &leq; 0.05.

@@

## Growth and morphology metrics
In this field study, colder temperatures and reductions in cumulative PPFD per day (Fig. `r airvarsfig`) most likely lead to the reduced growth in the free seedlings in the final weeks of the experiment (Fig. `r allomfig`).  Combined with severe growth reductions in the smallest container volumes the experiment was chosen to be harvested after 120 days. Over this duration height, diameter, and leaf area diverged between container volumes (Fig. `r allomfig`).  First, seedling leaf area significantly diverged between soil volumes (P < 0.026) during the 5^th^ week of the experiment. Following this period both height (8^th^} week) and then diameter (9^th^ week) significantly deviated across soil volumes (P < 0.002 & 0.001, respectively).  Negative growth effects then manifested as severely reduced height gain and declining leaf area through time with small soil volumes across the final two months of the experiment. Seedlings maintained diameter growth throughout the experiment, although marginal with smaller soil volumes in the final month. Final seedling height significantly increased with increasing soil volume (P < 0.001).  Increases in both final stem diameter (P < 0.001) and cumulative leaf area (both P < 0.001) were found with increasing soil volume and these differences were driven mainly by the largest container and the free seedling.

\
Total seedling biomass at harvest was significantly different across container volumes (P < 0.001) and with free seedlings (P < 0.001, Table. 1). We analyzed the relationship between biomass growth with each fold increase in soil volume and found an increase of 34 % with a doubling of pot size, consistent with the meta-analysis of Poorter et al. [-@poorter2012pot]. Additionally, plant biomass was highly correlated with total leaf area across all treatments (R^2^ = 0.97, P < 0.001). Differences in biomass partitioning to leaves, stems, and roots were not different across soil volumes after variation in seedling biomass within treatments was factored in the analysis (Fig.`r allocfig`a,b). Across all treatments, the final harvested root:shoot was conserved in these seedlings, with a slightly higher shoot than root mass on average ($\overline{x}$ = 0.904, Fig. `r allomfig`c).

\
Overall, SRL was higher in seedlings in containers compared to free seedlings but not significantly in every soil volume treatment   (Table. 1). Over the duration of the experiment SLA was higher in free seedlings but was not different across containers sizes (Table. 1, P < 0.001) and this pattern was evident in the first gas exchange measurement campaign (P < 0.001).

## Leaf chemistry
Leaf N % was significantly higher in free seedlings and the largest container volume at the onset of gas exchange measurements (6th week, P < 0.001).  Over the remaining duration of the experiment the smallest container volume had a significant reduction in leaf N % compared to other soil volumes, while free seedlings maintained the highest leaf N % (Table. 1, P < 0.001).  Leaf starch content in the smallest container was double that of free seedlings (P=0.039), while leaf soluble sugars did not differ across treatments throughout the experiment (Table. 1).  Differences in leaf starch between the free seedling and the smallest container were also evident during the first gas exchange campaign (P = 0.001).

## Gas exchange and photosynthetic parameters
A~sat~ and A~max~ were both significantly higher in the largest container volume and the free seedling at the first measurement campaign (both P < 0.001). Across all measurement campaigns A~sat~ (Fig. `r asatfig`) and A~max~ (Table 2.) were consistently higher in free seedlings than in containers (both P < 0.001). The interaction between photosynthetic capacity, leaf starch, and leaf N on a mass basis was marginally significant (P = 0.058) but A~max~ was highly correlated to both leaf N content and leaf starch (both P < 0.001). Across all measurement campaign A~max~ was higher when foliar N was also higher, usually associated with low levels of leaf starch (Fig. `r achemfig`a). A~max~  was also lower when leaf starch was high as higher leaf N often did not coincide with high leaf starch (Fig. `r achemfig`b)

\
The photosynthetic parameters J~max~ and Vc~max~ were not different within treatments at the beginning and end of gas exchange campaigns, therefore the parameter means per treatment are reported here (Table. 2).  Overall, both J~max~ and Vc~max~ were significantly higher in free seedlings with little variation between soil volume treatments (P = 0.001 & 0.002, respectively). Leaf dark respiration rates were not significantly different across soil volumes (Table. 2). The g~1~ parameter, generated for each seedling from the Medlyn et al [-@medlyn2012reconciling] optimal stomatal conductance model, was lowest in the free seedling and was marginally different across soil volume treatments (Table. 2). Predicted values of g~s~, using the g~1~ parameter, where highly correlated with observed values (R^2^ = .74, P < 0.001, data not shown).

\
Neither $\Psi$~pd~ nor $\Psi$~l~ were different across treatments, with mean values of -0.27 and -1.2 mPa across all seedlings, respectively. Although g~s~ in free seedlings was genearlly higher htan those in containers (Table 2. P=0.001), the mean rates for all seedlings were high at 0.37 mol H~2~0 m^-2^ s^-1^ and did not decline significantly across the experiment duration. Additionally, leaf $\delta$^13^C at final harvest was not different across treatments (Table. 1). Combined these indices provide strong evidence that water stress was not apparent on these well-watered seedlings throughout the experiment. Soil N % at harvest was not different across soil volumes ($\bar{x}$ = 04.5 %) and decreased approximately 3 % across all containers over the experiment duration. This indicates that nutrient leaching from free seedlings or from draining of containers following natural rainfall events did not differ between treatments.

@@

# Discussion
This study utilized a simple but novel field design to manipulate belowground sink limitation and physically restrict *Eucalyptus tereticornis* seedling production.  We then addressed questions regarding the coordination of A and growth by complementing empirical results with modelling approaches. We found that reductions in leaf A rates were not sufficient to explain observed reductions in total plant biomass production. We thus encourage the utilization of mass balance approaches and provide direction for future studies when testing factors that control plant growth under environmental change.

## Reductions in growth and physiology under sink limitation
As soon as seedlings became established both height and diameter growth were negatively affected by decreasing soil volume. This lead to the large reductions in harvested biomass in small containers when compared to free seedlings. These growth reductions were expected, as the impedance of root growth can cause reductions overall plant growth and activity [@mcconnaughay1991physical; @young1997mechanical]. It has been shown that roots undergoing difficult conditions may send inhibitory signals to the shoots that affect g~s~, cell expansion, cell division and the rate of leaf appearance [@passioura2002soil].  Here, this was evident in a large divergence in leaf area between seedlings in containers and free seedlings through time, with the eventual cessation of new leaf growth in seedlings in small containers.

\
Decreases in A~sat~ paralleled the reductions in allometric growth parameters of seedlings in containers. This initially suggests a strong link between growth and an apparent down regulation of A. However, there are several possible mechanisms that can explain reduced A in small pots including nutrient content, water or reduced sink strength [@poorter2012pot].  It was therefore necessary to examine each of these factors to distinguish if the induced belowground sink limitation actually triggered photosynthetic down regulation.

\
With high rates of g~s~, non-limiting leaf water potential and consistent leaf $\delta$^13^C across soil volume treatments there was little evidence of water stress inducing a limitation to A. This finding is consistent with other container size studies without drought treatments.  For example, reduced A~max~ in cotton seedlings grown at elevated CO~2~ was attributed to sink-limited feedback inhibition from inadequate rooting volume, not decreased g~s~ [@thomas1991root]. Additionally, severe reductions in A in coffee plants was not attributed to impacts of container size on leaf water potentials or g~s~ [@ronchi2006growth]. Here, it is likely that reductions in A of these well watered seedlings was instead the result of limiting soil nutrients on belowground sink strength.

\
Although the soil N pool declined evenly across all treatments leaf N was lowest in the smallest containers, suggesting that sink limitation was the greatest in these containers. This makes sense as small containers may reduce N uptake, either from physical root restriction or decreased supply, which affects growth, Rubisco limitation, sugar metabolism, and carbohydrate partitioning between source and sink tissues [@stitt1991rising; @hermans2006plants]. Mycorrhizal colonization could also have been affected in containers. Whether unrestricted mycorrhizal recruitment facilitated the increase in leaf N uptake in free seedlings is unknown. Regardless of whether sink limitation occurred from root or mycorrhizal restriction, the stark contrast in leaf N between free seedlings and seedlings in containers was significantly correlated to reductions in A.

\
As both rooting space and resources were finite in containers, the inability of seedlings to increase the capacity of the belowground C sink resulted in the buildup of C assimilate in leaves. The feedback inhibition of A from starch accumulation has been proposed, yet it is still not known whether there is a starch threshold that triggers the down-regulation process (Nebauer et al. 2011).  Here, declines in A~max~ where correlated with higher starch content throughout the experiment.  This agrees with findings from Equiza et a. (2006) where photosynthetic downregulation from reduced sink strength showed a greater correlation with starch than with sugar content in a deciduous conifer.  As starch content in the smallest containers was nearly double that of free seedlings this suggests the response of A to sink inhibition was regulated by this accumulation, as hypothesized.
